Adaptive evolution with aneuploidy and mutation

### 2022-03-07 ms/ms.tex
YR revised analysis up to line 541

% Fig single-locus model with and without aneuploidy
%% generated with with-aneuploidy_vs_no_aneuploidy.ipynb
\begin{figure}[h!]
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/fixation-plot-a.pdf}      
      \label{fig:fit}
  \end{subfigure}
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/fixation-plot-b.pdf}      
      \label{fig:fit-cumulative}
  \end{subfigure}
  \caption{
    \textbf{Single-locus model fit with and without aneuploidy.}
    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green).
In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450. Thus, the blue distribution is a better fit compared to the green, and the yellow histogram has a very poor fit.
The last bin contains all the simulations with time equal or greater than 3,000.
    \textbf{(B)}
	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A. The MAP likelihood (\Cref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
     }
  \label{fig:2n*-fixation}
\end{figure}

\begin{figure}[h!]
  \centering
\includegraphics[width=0.8\textwidth]{../figures/dynamics.pdf}
  \caption{
  \textbf{Posterior genotype frequency dynamics for the single-locus model.} 
  The posterior prediction for the frequencies of the four genotypes over time. Each of the 20 curves is the average of 10,000 simulations of the single-locus model using parameters drawn from the posterior distribution.
  }
  \label{fig:ppc-plot}
\end{figure}

% Fig tau plots
%% generated by diff-tau-plots.ipynb
\begin{figure}[h!]
  \centering
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-a.pdf}      
  \end{subfigure}
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-b.pdf}      
  \end{subfigure}
   \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-c.pdf}      
  \end{subfigure}
   \caption{
    \textbf{Genotype fixations for single-locus models with different $\tau$.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells. We then generated 10,000 simulations using the MAP estimate of each model and evaluated the fraction of simulations in which the genotype \euwt\ (\textbf{A}), \anwt\ (\textbf{B}), and \eumt\ (\textbf{C}) is fixed (y-axis) at each generation (x-axis). Note that \anmt\ did not fix. We can see that $\tau=100$ can be distinguished if time to loss of \euwt\ is known (panel~A) or if time to fixation or loss of \anwt\ is known (panel~B). It is harder to distinguish between $1\leq \tau \leq10$.
  }
  \label{fig:tau-plots}
  % TODO IK axis labels like in figure 3, also somehow the graphics looks different from previous figures
  % fixed means 100% ?

%%%%%%%%%%
\subsection*{Analysis}

Here, we aim to approximate the adaptation time (i.e., time to fixation of \eumt) in the single-locus model and how it is affected by the model parameters. As a stochastic model, there are two events that can occur in the \euwt\ population. The first is a mutation that produces a \eumt\ cell. The second is a missegregation event that produces a \anwt\ cell. The probability with which these events occur is significantly different. We are interested in the average waiting time until one of the two events occurs and then lead to adaptation (in the second case, it will be followed by a mutation to \anmt\ and loss of aneuploidy to \eumt).
We assume the waiting time for appearance of \eumt\ from \euwt\ and appearance of \anwt\ from \euwt are geometrically distributed with rates $p_{2n^*}$ and $p_{2n+1}$, respectively, where
\begin{subequations}
\begin{align}
p_{2n^*} &= 1-(1-2\mu \Delta w_{2n^*})^N \;, \\
p_{2n+1} &= 1-(1-2\delta\Delta w_{2n+1})^N \;,
\end{align}
\end{subequations}
where $\Delta w_i$ denotes the relative difference between the fitness of genotype $i$ and the population mean fitness.
However, we have two competing events: if one event occurs (e.g., mutation) then the other (aneuploidy) is unlikely to occur,  we need to find the minimum of these waiting times. The minimum of two geometric random variables with parameters $p_{2n^*}$ and $p_{2n+1}$ is also a geometric random variable with parameter $1-(1-p_{2n^*})(1-p_{2n+1})$, which is the inverse of the expected waiting time until the first event occurs.
The probability that aneuploidy will be the first event to occur is
\begin{equation}
p_A = \frac{p_{2n+1}}{p_{2n^*}+p_{2n+1}} \;.
\end{equation}
We are interested in the expected time to fixation of \eumt, and the events leading to this fixation depend on the event that occurs first. If \eumt\ appears first and avoids extinction by drift, then it will fix. If \anwt\ appears first and avoids extinction, then several other events must occur for fixation of \eumt. Thus, in this aneuploid trajectory, we need to approximate the expected waiting time to reach \anmt\ and finally \eumt.
In general, we assume that the waiting time until a new genotype occurs is much longer than the time to its fixation. Nonetheless, these fixation times must be included in a rigorous analysis. There are two ways to compute the fixation times and they differ in the assumption of the underlying process.

A deterministic-process approach is given by the classical bi-allelic single-locus equation, which provides an approximation for the expected time for genotype $i$ to reach frequency $x$ when starting from frequency $1-x$ in a population of genotype $j$ is % TODO MP is this correct?
 % TODO MP ref for equation? Crow and Kimura?
\begin{equation}
\tilde T_{i \to j} = 2\frac{\log{\big(\frac{x}{1-x}N\big)}}{\log(w_i/w_j)} \;,
\end{equation}
A stochastic-process approach for approximating $\tilde T_i$ is given by the diffusion-equation approximation \citep[eq.~17]{Kimura1969a}. 
Thus, the expected adaptation time via the aneuploidy trajectory is given by
\begin{equation}\label{eq:T_aneu}
 T_A = \frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \frac{1}{p_{2n+1^*}} + \frac{1}{p_{2n^*}} + \tilde T_{2n \to 2n+1} + \tilde T_{2n+1 \to 2n+1^*} + \tilde T_{2n+1^* \to 2n^*} \;,
 \end{equation}
the expected adaptation time via the mutation trajectory is given by 
\begin{equation}\label{eq:T_mut}
T_{M}=\frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \tilde T_{2n \to 2n^*} \;,
\end{equation}
and the expected adaptation time is given by
 \begin{equation}\label{eq:T_total}
 T = p_A T_{A}+(1-p_A) T_{M}.
 \end{equation}

@@

To validate the approximation in \Cref{eq:T_total}, we performed simulations of the full model (\Cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single})  without aneuploidy, $\delta=0$ (\Cref{fig:only mut}), and with the other parameters as inferred above. The approximation provides a good fit to simulation results except when the effective population size is larger than $5\cdot 10^6$ cells. In that case, the expected number of new mutants is ~0.5, and therefore we can expect that in many generations the fixation event will start with more than a single \eumt\ cell appearing in the same generation, which is in contrast to our assumption that an adaptation event starts from a single mutant cell.

 \begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{../figures/figgrid_noaneu.pdf}
 \caption{\textbf{Expected adaptation time without aneuploidy.}
  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B). Mean and 95\% CI from simulations in blue. Analytic approximation (\Cref{eq:T_total}) in orange. Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$. In panel A, $\mu=9.6\cdot 10^{-8}$ and in panel B, $N=6\cdot 10^6$.
  }
  %TODO MP add panel labels A and B
  % The error bars a quite large, can you make them smaller by running more simulations?
  % Add more x ticks for panel A (10^6, 5 10 ^6 ...)
  % maybe start x axis in panel A with lower population size? 10^4 or 10^5 ? does the approximation work well there?
 \label{fig:only mut}
 \end{figure}

% Discussion
\section*{Discussion}

>>>>

% Fig single-locus model with and without aneuploidy
%% generated with with-aneuploidy_vs_no_aneuploidy.ipynb
\begin{figure}[p]
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/fixation-plot-a.pdf}      
      \label{fig:fit}
  \end{subfigure}
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/fixation-plot-b.pdf}      
      \label{fig:fit-cumulative}
  \end{subfigure}
  \caption{
    \textbf{Single-locus model fit with and without aneuploidy.}
    \textbf{(A)} The distribution of time to fixation of \eumt\ (i.e., adaptation time) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange), and a model fitted to the data assuming $\delta=0$ (green).
In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and another by generation 2,350 (dashed lines) but not before generation 450. Thus, the blue distribution is a better fit compared to the green, and the yellow histogram has a very poor fit.
The last bin contains all the simulations with time equal or greater than 3,000.
    \textbf{(B)}
	Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy in blue and green, respectively, and corresponding to the blue and green bars in panel A. The MAP likelihood (\Cref{eq:heatstress-likelihood}) is 0.84 and 0.67 with and without aneuploidy, respectively.
     }
  \label{fig:2n*-fixation}
\end{figure}

\begin{figure}[p]
  \centering
\includegraphics[width=0.8\textwidth]{../figures/dynamics.pdf}
  \caption{
  \textbf{Posterior genotype frequency dynamics for the single-locus model.} 
  The posterior prediction for the frequencies of the four genotypes over time. Each of the 20 curves is the average of 10,000 simulations of the single-locus model using parameters drawn from the posterior distribution.
  }
  \label{fig:ppc-plot}
\end{figure}

% Fig tau plots
%% generated by diff-tau-plots.ipynb
\begin{figure}[p]
  \centering
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-a.pdf}      
  \end{subfigure}
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-b.pdf}      
  \end{subfigure}
   \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-diff-c.pdf}      
  \end{subfigure}
   \caption{
    \textbf{Genotype fixations for single-locus models with different $\tau$.} We estimated the parameters for different models, each assuming a different value of $\tau$, the fold-increase in mutation rate in aneuploid cells. We then generated 10,000 simulations using the MAP estimate of each model and evaluated the fraction of simulations in which the genotype \euwt\ (\textbf{A}), \anwt\ (\textbf{B}), and \eumt\ (\textbf{C}) is fixed (y-axis) at each generation (x-axis). Note that \anmt\ did not fix. We can see that $\tau=100$ can be distinguished if time to loss of \euwt\ is known (panel~A) or if time to fixation or loss of \anwt\ is known (panel~B). It is harder to distinguish between $1\leq \tau \leq10$.
  }
  \label{fig:tau-plots}
  % TODO IK axis labels like in figure 3, also somehow the graphics looks different from previous figures
  % fixed means 100% ?
\end{figure}

\pagebreak
%%%%%%%%%%
\subsection*{Analysis}

Here, we aim to approximate the adaptation time (i.e., time to fixation of \eumt) and analyse how it is affected by the model parameters.
As a stochastic model, there are two events that can occur on the \euwt\ genotype background. The first is a mutation that produces a \eumt\ genotype. The second is a missegregation event that produces a \anwt\ genotype. The probabilities with which these events occur is significantly different. We are interested in the average waiting time until one of these two genotype appears and then leads to adaptation, i.e., appearance and fixation of the \eumt; in the second case, appearance of \anwt\ will be followed by a mutation to \anmt\ and loss of aneuploidy to \eumt.

We assume the waiting time for appearance of \eumt\ from \euwt\ and appearance of \anwt\ from \euwt\ are geometrically distributed with rates $p_{2n^*}$ and $p_{2n+1}$, respectively, where
\begin{subequations}
\begin{align}
p_{2n^*} &= 1-(1-2\mu \Delta w_{2n^*})^N \;, \\
p_{2n+1} &= 1-(1-2\delta\Delta w_{2n+1})^N \;,
\end{align}
\end{subequations}
where $\Delta w_i$ denotes the relative difference between the fitness $w_i$ of genotype $i$ and the population mean fitness, $\bar{w}$.

However, we have two competing events: if one event occurs (e.g., mutation), then the other event (e.g., aneuploidy) is unlikely to occur. Thus, we need to find the minimum of these waiting times. The minimum of two geometric random variables with rates $p_{2n^*}$ and $p_{2n+1}$ is also a geometric random variable with rate $1-(1-p_{2n^*})(1-p_{2n+1})$, which is the inverse of the expected waiting time until the first event occurs.

% TODO MP: this seems like repetition of things written above, right? should we remove this? or if there is something important here, merge it to previous paragraphs?
%We are interested in the expected time to fixation of \eumt, and the trajectory leading to this fixation depends on the event that occurs first. If \eumt\ appears first and avoids extinction by drift, then it will fix. If \anwt\ appears first and avoids extinction, then several other events must occur for fixation of \eumt. Thus, in this aneuploid trajectory, we need to approximate the expected waiting time to reach \anmt\ and finally \eumt.

Given a mutation occurs first, % TODO MP: I added this sentence, is that right?
the expected adaptation time via the mutation trajectory is 
\begin{equation}\label{eq:T_mut}
T_{M}=\frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + T_{2n \to 2n^*} \;,
\end{equation}
where $T_{i \to j}$ is the waiting time for fixation of genotype $i$ in a population of genotype $j$, given it starts with a single copy and does not go to extinction by drift.
Given that aneuploidy occurs first, % TODO MP: I added this sentence, is that right?
the expected adaptation time via the aneuploidy trajectory is approximated by
\begin{equation}\label{eq:T_aneu} 
% TODO MP: p_{2n+1^*} is not defined, and why is there 1/p_{2n^*} here? shouldn't it have the waiting time for 2n+1* to appear on 2n+1 and then waiting time for 2n* to appear on 2n+1*? something is wrong here.
 T_A = \frac{1}{1-(1-p_{2n^*})(1-p_{2n+1})} + \frac{1}{p_{2n+1^*}} + \frac{1}{p_{2n^*}} + T_{2n \to 2n+1} + T_{2n+1 \to 2n+1^*} + T_{2n+1^* \to 2n^*} \;.
 \end{equation} % TODO MP do we need the tilde ?
Finally, the expected adaptation time is 
 \begin{equation}\label{eq:T_total}
 T = p_A T_{A}+(1-p_A) T_{M} \;,
 \end{equation}
 where $p_A$ is he probability that aneuploidy will be the first event to occur,
\begin{equation}
p_A = \frac{p_{2n+1}}{p_{2n^*}+p_{2n+1}} \;.
\end{equation}

We assume that the waiting time until a new genotype occurs is much longer than the waiting time for its fixation in the population. Nevertheless, these fixation times must be included in a rigorous analysis. There are two ways to compute the fixation times and they differ in the assumption of the underlying process.
A deterministic-process approach is given by the classical bi-allelic single-locus equation.
It approximates the expected time for genotype $i$ to reach frequency $x$ when starting from frequency $1-x$ in a population of genotype $j$,
 % TODO MP ref for equation? Crow and Kimura? YR: yes
\begin{equation} \label{eq:T_fix_x}
 T_{i \to j} \approx 2\frac{\log{\big(\frac{x}{1-x}N\big)}}{\log(w_i/w_j)} \;,
\end{equation} 
A stochastic-process approach for approximating $T_{i \to j}$ is given by the diffusion-equation approximation \citep[eq.~17]{Kimura1969a}.

@@

First, to validate the approximation in \Cref{eq:T_total}, we performed simulations of the full model (\Cref{eq:mutation-aneuploidy-single,eq:drift-single,eq:selection-single}) without aneuploidy, i.e., $\delta=0$ (\Cref{fig:only mut}), and with the other parameters as inferred above. The approximation provides a good fit to simulation results except when the effective population size is larger than $5\cdot 10^6$ cells. In that case, the expected number of new mutants per generation is roughly 0.5, and  we can expect that in many generations the fixation event will start with more than a single \eumt\ cell appearing in the same generation, which violates our assumption that an adaptation event starts from a single mutant cell.

 \begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{../figures/figgrid_noaneu.pdf}
 \caption{\textbf{Expected adaptation time without aneuploidy.}
  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (\textbf{A}) or as a function of the mutation rate, $\mu$ (\textbf{B}). Mean and 95\% CI from simulations in blue. Analytic approximation (\Cref{eq:T_total}) in orange. Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$. In panel A, $\mu=9.6\cdot 10^{-8}$. In panel B, $N=6\cdot 10^6$.
  }
  %TODO MP 
  % which approximation do you use here for T_i->j ?
  % The error bars a quite large, can you make them smaller by running more simulations?
  % --> or by making the y axis in log scale!!
 \label{fig:only mut}
 \end{figure}

 \begin{figure}[h!]
 %\centering
 \includegraphics[width=1\textwidth]{../figures/figgrid_nodirectmut_ylog_uncorr0.jpg}
 \caption{\textbf{Expected adaptation time without mutations from $\euwt$ to $\eumt$.}
  The expected time to appearance and fixation of \eumt\ as a function of the population size, $N$ (\textbf{A}) and as a function of the mutation rate, $\mu$ (\textbf{B}) and as a function of the aneuploidy rate, $\delta$ ({C}). Blue error bars for mean and 95\% CI from simulations. Green line for the analytic approximation (\Cref{eq:T_total}). % TODO why green? in previous figure it was orange.
  Red curve for the approximation with $T_{\euwt \to \anwt}=T_{\anwt \to \anmt}=0$. 
  Orange curve for the approximation with $x=0.017$ in \Cref{eq:T_fix_x} for the fixation time of $\eumt$; we found this gives a good fit in panel B.
  Here, $w_\anwt=1.021,\, w_\anmt=1.025$ and $w_\eumt=1.028$, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$ unless parameter is varied on the x-axis.
  }
 \label{fig:no direct mut}
 \end{figure}

Second, we performed simulations without mutations from $\euwt$ to $\eumt$ (\Cref{fig:no direct mut}). 
% TODO MP: I comment this out - not clear what "Before" refers to, and not sure we need this sentence. what do you think?
%Before, we had the optimal scenario of waiting until the first successful mutation appears and then waiting until is reaches fixation. Our approximations did quite well there. 
Here, aneuploidy is transient and fixation of \anwt\ or \anmt\ could be outpaced by the arrival of the next genotype. This would explain the discrepancy between the green curves in \Cref{fig:no direct mut}.
To asses this further, \Cref{fig:fixT vs waitT} provides a clear picture. At least for intermediate to high values of the respective range of the parameters, the waiting time until a succesful new type appears is far smaller than that to fixation of the old type. Taking this into account, it seems reasonable to ignore the fixation time of $\anwt$ and $\anmt$, which results in the better fitting red curves in \Cref{fig:no direct mut}. Inferring $x=0.017$ for $\tilde{T}_{\anmt \rightarrow \eumt}$ to get the best fit in panel B, does not imply that it is also a best fit for panels A and C. However, we stop simulations in state $\eumt$ if they have frequency $0.95$.

  The discrepancies in \Cref{fig:no direct mut}C seem to be particularly odd as the approximations are nearly constant and the simulations decreas log-linearly with increasing $\delta$.

 \begin{figure}[h!]
 %\centering
 \includegraphics[width=1\textwidth]{../figures/figgrid_fixT_vs_waitT.jpg}
 \caption{\textbf{Expected time of fixation and waiting time events on the aneuploid trajectory.}
  The expected time to appearance and fixation of each step seperately in the aneuploid trajectory as a function of the population size, $N$ (panel A) and as a function of the mutation rate, $\mu$ (panel B) and as a function of the aneuploidy rate, $\delta$ (panel C). Solid curves give the fixation time and dashed curves the waiting time. Curves of the same color indicate the same base population state;  e.g. the blue dashed curve is the waiting time in state $\anwt$ until a succesful individual of type $\anmt$ appears. The corresponding solid curve indicates the time to fixation of the population in state $\anwt$. Here, $w_\anwt=1.021,\,   w_\anmt=1.025$ and $w_\eumt=1.028$. Parameter values in panels where it is not varied, $N=6\cdot 10^6$,\, $\mu=9.6\cdot 10^{-8}$ and $\delta=5.4\cdot 10^{-4}$.  }
 \label{fig:fixT vs waitT}
 \end{figure}

In summary, whereas the two possible trajectories to $\eumt$ do not compete against each other in both \Cref{fig:only mut} and \Cref{fig:no direct mut}, the approximations are doing well only in restricted parameter ranges. These tend to be, where the respective parameters are rather small. If they grow bigger, they are violating some assumptions of our approximations.

%%%%%%%%%%
% Discussion
\section*{Discussion}
